Chaotic mixing of tidal debris

### 2015-06-27 ms.tex
more refs

The morphological (density) evolution of the tidal debris therefore depends only on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By modeling the observed phase-space density of stream stars along with the host galaxy potential it is hoped that we may infer the 3D mass distribution of the host. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies; the known streams span a large range of distances --- from $\approx10$ to $100~{\rm kpc}$ --- and progenitor masses --- from $\approx$10$^3$ to $\approx$10$^8~\msun$ in stellar mass --- \citep[][]{cite many}. There has been extensive work on developing methods to use data from these streams to measure properties of Milky Way's dark matter halo. These methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to N-body simulations \citep[see, e.g., the introduction of][]{kuepper15}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms. 
% internal question: This is all just about the methods -- do I need sentences or a paragraph on results?

@@

\item \emph{Substructure}: Cosmological simulations predict that dark matter halos are filled with substructure in the form of dark matter subhalos. If they exist, these subhalos may account for up to $\approx$1-10\% of the mass of the dark matter \citep[e.g.,][]{diemand07} and therefore may contribute significantly to and orbit in the large-scale potential of any galaxy. Gravitational scattering due to subhalo interactions has been studied, however in the halos of galaxies where dynamical times are long, the scattering cross-section for \emph{strong} encounters is very small \citep{??}. Instead, the collective effect of the subhalos may instead act as a noise term in the Hamiltonian of any halo orbit. This subhalo-induced colored noise --- which depends on the mass spectrum and distribution of subhalos --- may also act to simply increase the magnitude of chaos along orbits, and destabilize sufficiently non-resonant orbits \citep[see, e.g.,][]{kandrup00}.

@@

Tidal stream modeling is one of the most promising ways to constrain the 3D mass distribution around the Milky Way at distances of 10s to 100s of kpc. Methods that use tidal streams to infer properties of the Galactic potential typically operate by constructing models of the debris distribution using either the present-day phase-space density \citep[e.g.,][]{kuepper12, kuepper15, amorisco and others}, time-of-disruption phase-space density \citep{apw13, apw14}, or the density in angle-action coordinates \citep{sanders14, bovy14}. All of these methods may fail or produce uninterpretable results if modeling globular-cluster streams on mildly chaotic orbits. For each of these methods, it is important to understand the failures and biases introduced by ignoring chaotic orbit evolution.

@@

Our main conclusions are summarized as follows:
\begin{enumerate}
	\item the Lyapunov time and frequency diffusion time are powerful indicators of chaos, but do not capture the importance of small-scale chaotic diffusion for the density evolution of small ensembles of orbits (tidal debris);
	\item tidal debris becomes diffuse and thus harder to observe on weakly chaotic orbits when the small-scale chaotic diffusion of the fundamental frequencies has a scale comparable to the internal spread in frequencies of the debris;
	\item the details of the enhanced mixing along weakly chaotic orbits depends on the resonant structure of the potential;
	\item the variance of the fundamental frequencies of an orbit over a given time window is a strong predictor of the importance of small-scale chaotic frequency diffusion on the resulting morphology of tidal debris.
\end{enumerate}

>>>>

The morphological (density) evolution of the tidal debris therefore depends only on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also determined by the shape and radial profile of the gravitational potential of the host galaxy. By modeling the observed phase-space density of stream stars along with the host galaxy potential it is hoped that we may infer the 3D mass distribution of the host. Many tidal streams are observed around the Milky Way, M31, and other nearby galaxies; the known streams span a large range of distances --- from $\approx10$ to $100~{\rm kpc}$ --- and progenitor masses --- from $\approx$10$^3$ to $\approx$10$^8~\msun$ in stellar mass --- \citep[][]{ibata94, belokurov06,grillmair06a,grillmair06b}. There has been extensive work on developing methods to use data from these streams to measure properties of Milky Way's dark matter halo. These methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to N-body simulations \citep[see, e.g., the introduction of][]{kuepper15}. All methods have been tested in some way on simulated observations of data and these tests typically demonstrate the recovery of parameters for analytic, static potential forms. 
% internal question: This is all just about the methods -- do I need sentences or a paragraph on results?

@@

\item \emph{Substructure}: Cosmological simulations predict that dark matter halos are filled with substructure in the form of dark matter subhalos. If they exist, these subhalos may account for up to $\approx$1-10\% of the mass of the dark matter \citep[e.g.,][]{diemand07} and therefore may contribute significantly to and orbit in the large-scale potential of any galaxy. Gravitational scattering due to subhalo interactions has been studied, however in the halos of galaxies where dynamical times are long, the scattering cross-section for \emph{strong} encounters is small. Instead, the collective effect of the subhalos may instead act as a noise term in the Hamiltonian of any halo orbit. This subhalo-induced colored noise --- which depends on the mass spectrum and distribution of subhalos --- may also act to simply increase the magnitude of chaos along orbits, and destabilize sufficiently non-resonant orbits \citep[see, e.g.,][]{kandrup00}.

@@

Tidal stream modeling is one of the most promising ways to constrain the 3D mass distribution around the Milky Way at distances of 10s to 100s of kpc. Methods that use tidal streams to infer properties of the Galactic potential typically operate by constructing models of the debris distribution using either the present-day phase-space density \citep[e.g.,][]{kuepper12, kuepper15, gibbons14}, time-of-disruption phase-space density \citep{apw13, apw14}, or the density in angle-action coordinates \citep{sanders14, bovy14}. All of these methods may fail or produce uninterpretable results if modeling globular-cluster streams on mildly chaotic orbits. For each of these methods, it is important to understand the failures and biases introduced by ignoring chaotic orbit evolution.

@@

Our main results and conclusions are summarized as follows:
\begin{enumerate}
	\item ``stream-fanning'' of tidal streams on weakly chaotic orbits \citep[as seen in simulations by][]{pearson15} occurs due to small-scale chaotic evolution of the fundamental frequencies of the debris star orbits;
	\item the Lyapunov time and frequency diffusion time are powerful indicators of chaos, but do not capture the importance of small-scale chaotic diffusion for the density evolution of small ensembles of orbits (tidal debris);
	\item tidal debris becomes diffuse and thus harder to observe on weakly chaotic orbits when the small-scale chaotic diffusion of the fundamental frequencies has a scale comparable to the internal spread in frequencies of the debris;
	\item the details of the enhanced mixing along weakly chaotic orbits depends on the resonant structure of the potential;
	\item the variance of the fundamental frequencies of an orbit over a given time window is a strong predictor of the importance of small-scale chaotic frequency diffusion on the resulting morphology of tidal debris.
\end{enumerate}
